### 2018-07-25 introduction.tex
Extend references

Novozymes A/S \ac{NZ} is a company whose line of business consists of the development of products performing chemical transformations in economically relevant processes. The application of these products, instead of conventional solutions, has the advantage of requiring less chemical substances, potentially simplifying industrial processes, reducing their costs and their environmental impact. Notorious examples of such applications include waste-water treatment, household care or fermentation. For example, a product called Fermax was recently released to prevent foam development during the sugarcane ethanol production (see figure \ref{fig:ethanol}).

@@

The company\textquotesingle s name is inspired by the molecular agents that power these transformations, \textbf{enzymes}. Enzymes consist of a extremely diverse family of molecules catalyzing a plethora of biochemical reactions happening in living beings, from bacteria to humans.

@@

The majority of enzymes consist of proteins. But what are proteins? Proteins are molecules made up of 20 basic units, called aminoacids. All aminoacids share a common chemical structure, where a carbon atom ($C_\alpha$) is covalently bonded to a hydrogen atom, a carboxyl group (\ce{COOH}), an amino group (\ce{NH_2}), and last but not least, a radical, also called side chain of the aminoacid. A slight deviation from this pattern exists in proline, where the radical is bound to the nitrogen atom, making it an iminoacid. The side chain differs between aminaocids and creates the aminoacidic diversity   (see figure \ref{fig:aminoacids}).

@@

The carboxyl group and amino group of different aminoacids can undergo a dehydration that generates a covalent bond, called peptidic bond, subsequently joining two aminoacids (see figure \ref{fig:peptidic_bond}). The remaining amino and carboxyl groups can in turn be joined to new aminoacids, giving way for the formation of a biopolymer, where each subunit is an aminoacid. A peptide consists of a small chain of aminoacids joined this way. If made long enough, it acquires complex structures and becomes a protein proper \cite{Nelson2008}.

@@

More concretely, proteomics studies how proteins are capable of coordinating their functions by means of complex systems of regulation.  Proteomics aims at unraveling the complexity, understanding how its powerful properties emerge from its individual components CITE NETWORK THEORY. For instance, the interplay of proteomics and biotechnology helps researchers learn which proteins are key in the onset of a frequent disturbance of these regulations: cancer \cite{Saraon2012}. The application of proteomics to biological problems is not restricted to medical problems like that, but also biotechnological processes like those mentioned above.

>>>>

Novozymes A/S \ac{NZ} is a company whose line of business consists of the development of products performing chemical transformations in economically relevant processes. The application of these products, instead of conventional solutions, has the advantage of requiring less chemical substances, potentially simplifying industrial processes, reducing their costs and their environmental impact \cite{madigan2000brock}. Notorious examples of such applications include waste-water treatment, household care or fermentation. For example, a product called Fermax was recently released to prevent foam development during the sugarcane ethanol production (see figure \ref{fig:ethanol}).

@@

The company\textquotesingle s name is inspired by the molecular agents that power these transformations, \textbf{enzymes}. Enzymes consist of a extremely diverse family of molecules catalyzing a plethora of biochemical reactions happening in living beings, from bacteria to humans \cite{Nelson2008}.

@@

The majority of enzymes consist of proteins. But what are proteins? Proteins are molecules made up of 20 basic units, called aminoacids. All aminoacids share a common chemical structure, where a carbon atom ($C_\alpha$) is covalently bonded to a hydrogen atom, a carboxyl group (\ce{-COOH}), an amino group (\ce{-NH_2}), and last but not least, a radical, also called side chain of the aminoacid. A slight deviation from this pattern exists in proline, where the radical is bound to the nitrogen atom, making it an iminoacid. The side chain differs between aminaocids and creates the aminoacidic diversity   (see figure \ref{fig:aminoacids}).

@@

The carboxyl group and amino group of different aminoacids can undergo a dehydration that generates a covalent bond, called peptidic bond, subsequently joining two aminoacids (see figure \ref{fig:peptidic_bond}). The remaining amino and carboxyl groups can in turn be joined to new aminoacids, giving way for the formation of a biopolymer, where each subunit is an aminoacid. A peptide consists of a small chain of aminoacids joined this way. If made long enough, it acquires complex structures and becomes a protein \cite{Nelson2008}.

@@

More concretely, proteomics studies how proteins are capable of coordinating their functions by means of complex systems of regulation.  Proteomics aims at unraveling the complexity, understanding how its powerful properties emerge from its individual components \cite{Milo2002}. For instance, the interplay of proteomics and biotechnology helps researchers learn which proteins are key in the onset of a frequent disturbance of these regulations: cancer \cite{Saraon2012}. The application of proteomics to biological problems is not restricted to medical problems such as cancer, but also biotechnological processes like those mentioned above.

### 2018-07-25 msreview.tex
Add Kruschke-like diagram and fix some section referencing
Extend references

The main source of data in proteomics is \ac{MS}. However, different approaches to how these technique is used make two paradigms in proteomics analysis: top-down and bottom-up. In the top-down paradigm, intact proteins are directly used for the analysis. In the bottom-up paradigm (see figure \ref{fig:proteomics_overview}), the proteins are first cleaved into smaller parts, and these parts are then used for identification, characterization, and quantification. These smaller parts are peptides \cite{Barsnes2008}, consisting of around 10 chained aminoacids.  Such peptides acquire physicochemical properties fitting the requirements of the downstream analytical methods, mainly the mass spectrometer (MS), which performs the data acquisition. The bottom-up paradigm is most often used because peptides are much more suitable to analysis by mass spectrometry, as illustrated in section \ref{subsec:the_detector}. This review will focus on the data-dependent approach within bottom-up proteomics, which currently is the most frequent workflow in proteomics \cite{Sinitcyn2018}. In this approach, the equipment is configured for the analysis of one peptide at a time, with the drawback of being biased for the most abundant peptides.

@@

An \ac{MS} experiment starts with the generation of a protein sample from the biological system of interest. Proteins are then denaturalized  so as to remove bias due to the divergent properties acquired by folded proteins. Then, proteins are subjected to digestion with specific proteases i.e protein-cutting molecules, which cut the aminoacidic chains following a predictable pattern. Trypsin is the most frequently used protease. It cuts peptidic bonds whenever a positively charged residue, either Lysine (K) or Arginine (R), lies on the carboxyl side of the peptidic bond. Since roughly 1/10 residues are either R or K, the average peptide length is 10 residues, as mentioned before. As demonstrated in \ref{subsec:the_detector}, this length distribution is fitted to the resolution of the MS analyzer. Moreover, in as much R and K are positively charged aminoacids (see figure \ref{fig:aminoacids}), the resulting peptide is guaranteed in most cases to be able to capture at least one charge, which is key in the \ac{MS} workflow as described in section \ref{sec:the_mass_spectrometer}. All of these properties combined, together with its low prize, makes Trypsin the protease of choice in this step for most cases.

@@

Missed cleavages can happen due to steric impediments or the presence of specific aminoacids that can weaken the enzyme\textquotesingle s function. This is the case of Trypsin whenever the residue on the other side of the peptidic bond is Proline. Altogether, a variability is created in the cleavage process that, though limited, needs to be taken care of in downstream analysis, as it could introduce biases in peptide observability.

@@

If presented with the problem of analyzing a mixture of peptides, the capacities of mass spectrometers are easily overwhelmed by a too complex mixture, resulting in the analysis of only a minor part of the total protein of the sample. This can be surmounted by analyzing one peptide in the sample at a time. The required sample separation is achieved by High-Performance Liquid Cromatography (\ac{HPLC}) methods, like reverse phase chromatography (separating on hydrophobicity) and strong cation exchange chromatography (separating on isoelectric point) \cite{Barsnes2008}.

%CITE 4.2 computational methods.
During \ac{HPLC}, the peptide mix is loaded into a column containing a stationary and a solid phase. These phases create an environment where peptides interact differently based on their physico-chemical properties, set by the nature of the phases. The output of the column, called elute, will consist of subsets or fractions of peptides leaving the column at different retention times (\ac{RT}) i.e the amount of time passed before the peptide is observed in the mass spectrometer. Therefore, the input to the spectrometer will consist of one peptide at any given time.

@@

The mass spectrometer is the ensemble of pieces of equipment analyzing a peptide like as those generated following the workflow enunciated in sections \ref{sec:sample_processing} and \ref{subsec:peptide_separation}. It consists of three main parts: an ion source, a mass analyzer, and a detector (see figure \ref{fig:mass_spectrometer}).

@@

This step is performed in the ion source \cite{Barsnes2008}. The most frequent ionization methods in proteomics are Matrix-Assisted Laser Desorption-Ionization (\ac{MALDI}) and Electro Spray Ionization (\ac{ESI}) \cite{Mirzaei2016}. Most peptides ionized by MALDI will acquire a single charge, whereas ESI can provide multiple charges (+2, +3, etc) too. Thus, the charge exhibited by an ion is not obvious when produced via ESI. Moreover, ESI can be run online with the right robotic equipment, while MALDI demands waiting time for vacuum generation. Finally, due to the chemical nature of the matrix components, MALDI ionizes more easily peptides containing aminoacids featuring aromatic rings (PYW), thus introducing a bias. Bias in ESI is less predictable. All the sources of bias introduced during ionization are grouped into the competitive ionization problem \cite{Tang2004}.

The acquired charge yields a mass/charge (\ac{m/z}) ratio, a property that is applied in the downstream component separation and measurement steps.

@@

On the other hand, due to the natural occurrence of isotopes, particularly \ce{^{13}_{}C}, the same peptide will induce the measurement of several signals with very close \ac{m/z} values. They constitute the isotopic envelope of the ion (see figure \ref{fig:envelope}), and represent the signal created by peptides containing an increasing number of \ce{^{13}_{}C} atoms. Every time a \ce{^{12}_{}C} is replaced by \ce{^{13}_{}C}, the mass increases by 1 Da. Even though the natural abundance of \ce{^{13}_{}C} is 1.1 \%, the sheer number of carbon atoms in a peptide makes it likely that at least one or even more carbon atoms will be \ce{^{13}_{}C}, eventually driving the pure \ce{^{12}_{}C} signal to comparatively small intensity values, and down to intensities below the background noise. Such event can be problematic if it entails that the \ce{^{13}_{}C} peak is confused for the \ce{^{12}_{}C} peak.

@@

The resolution achieved by modern equipment allows for the distinction of each individual signal in most isotopic envelopes. Remarkably, the separation across peaks in the envelope can be used to infer the charge of the peptide, as increases of 1 Da at charge 1 will induce a separation of 1 \ac{m/z}, while at charge 2 it will be $1/2 = 0.5$ \ac{m/z}, at 3 $1/3 = 0.33$ \ac{m/z}, and so on (see figure \ref{fig:envelope}).

@@

Shotgun proteomics analyses make use of two or more mass spectrometers connected in series, giving rise to the so-called Tandem MS (\ac{MS/MS}) workflow. In this setting, each mass spectrometer collects a different type of spectra and thus different information (see figure \ref{fig:shotgun}. An extra spectrometer, usually a quadrupole, is introduced in between.

@@

\item The first spectrometer records the intensity versus \ac{m/z} ratio of the peptides eluting from the column at a given time and is used to filter ions exhibiting a selected \ac{m/z} ratio.

@@

\item The last spectrometer records the intensity versus \ac{m/z} ratio of the fragments produced in the previous step. The produced spectrum can be use to read out the peptide sequence.

@@

The information that can be extracted from \ac{MS1} scans consists of the \ac{m/z} ratio and retention time of the peptide ion, but not its sequence. Having the latter is paramount if the spectrum is to be matched to a theoretical peptide.

@@

Fortunately, the peptide sequence of the ion can be inferred if fragmentation is performed. During peptide fragmentation, bonds along the peptidic chain are broken, turning the peptide into smaller fragments. These fragments will consist of truncated versions of the original peptide at different positions, thus making it possible to read an \ac{m/z} ratio difference between any pair of fragment ions. The difference can be exploited to deduce which aminoacid makes up for that difference. If this process is repeated for enough pairs of contiguous fragments, with the right software, a sequence can be read from the \ac{MS2} spectrum, as  explained in section \ref{sec:search_engines}.

@@

Computational analysis of MS data starts with the matching of the spectra to a referene proteome (see figure \ref{fig:computational_analysis}). \ac{MS} search engines are capable of performing the crucial step of peptide to spectrum matching (PSM). During this step, search engines perform \textit{in silico} prediction of the peptides produced in the protein digestion step and their expected spectrum.

@@

Given the stochastic nature of the protein cleavage and spectra recording processes, the resulting data exhibit variability manifested in both missing and spurious peaks. Furthermore, random (wrong) matches can be returned by the PSM process when running against a sufficiently big database. This translates to the generation of multiple matches, of which one, if any, will be correct. Therefore, the lists of matches need to be somehow ranked by goodness-of-fit. The issue is addressed by search engines through the deployment of statistical models that provide scoring systems. Assuming the correct protein is present in the database, a good scoring system should give the best score to the right peptide. Under this circumstances, if repeated for several peptides, enough evidence for the presence of individual proteins can be collected.

Multiple search engines exist that implement different matching and scoring algorithms. The most modern ones include MS-GF+, MS-Amanda, Comet, X!Tandem or Andromeda. Notably, the results of each individual search engine can be combined to gather their strengths, at the expense of an increased computational cost and time \cite{Shteynberg2013}.

@@

A common filter is the false discovery rate (FDR), usually set to 1\%, indicating that after the its application, only one out of a hundred filtered matches are expected to be false positives (wrong matches).

The most commonly used method to compute the \ac{FDR} of a list of matches is the target-decoy search. Using this method, the search engine replicates the matching process, using the same spectra, but instead against a decoy database. The decoy database is generated by reversing or more generally applying a randomization technique upon the sequences present in the original database (target).

@@

The minimal FDR at which a given PSM is considered a valid match constitutes the PSM\textquotesingle s q-value, i.e it is the smallest FDR we can allow while still keeping the PSM. Related to the q-value, the Posterior Error Probability (PEP) is an estimate of the probability of a given PSM of being an incorrect assignment (see figure \ref{fig:pep}). The PSM\textquotesingle s confidence is just defined as $1-PEP$ \cite{Nesvizhskii2010}. PEP can be computed from the decoy search results and provides another useful measurement of the uncertainty in the target results.

@@

The combination of all the aforementioned computational analyses yields a list of protein identities that reports the protein composition i.e qualitative information of the original sample. However, in most proteomics applications, quantitative data can be of great interest, as many biological phenomena are manifested mainly through changes in the protein abundances, rather than protein presence alone. For instance, cancer cells in response to a drug could modulate the abundance of several proteins without removing them from the cytosol or introducing new ones.

@@

Two paradigms exist in protein quantification: label-based and label-free. In label-based quantification, originally identical peptides from a number of different samples are made distinguishable by their masses via the incorporation of a label. All label-based methods simultaneously analyze several samples in each experiment, removing the difficulties associated with between-run variability \cite{Barsnes2008}. The finite number of "plexes" available for a given label sets the limit to how many samples can be differentially quantified \cite{Cox2014}. Different techniques, like Stable Isotope Labeling by Amino acids in Cell culture (SILAC) or Isotope-Coded Affinity Tags (ICAT), differ in the nature of the label and the way it is introduced. Remarkably, peptide labeling costs can be quite high. This, together with the limited amount of samples that can be compared makes a case for label-free quantification.

@@

On the one hand, spectrum counting based quantification is the simplest method in proteomics. The number of peptides a peptide is detected in the spectrometer is detected as proxy for its abundance. It relies on the rationale that highly abundant peptides will have a higher intensity and are thus more likely to trigger the acquisition of MS/MS spectra. These methods have the advantage that they are very simple to implement and don't require any further data processing.

@@

For a more robust XIC based quantification, a feature extraction step is usually executed to extract the apex intensity of the identified peak clusters, as shown in figure \ref{figure:moff_apex}.

@@

It is for this reason that peptide-based models offer the statistical framework required to learn as much from the data as possible. This translates into improved results when compared to the other aforementioned methods \cite{Goeminne2015}. This hypothesis is the motivation for the method explained in chapter \ref{chap:model}.

>>>>

The main source of data in proteomics is \ac{MS}. However, different approaches to how this technique is used make for two paradigms in proteomics analysis: top-down and bottom-up \cite{Joshi2017}. In the top-down paradigm, intact proteins are directly used for the analysis \cite{Sinitcyn2018}. In the bottom-up paradigm (see figure \ref{fig:proteomics_overview}), the proteins are first cleaved into smaller parts, and these parts are then used for identification, characterization, and quantification. These smaller parts are peptides \cite{Barsnes2008}, consisting of around 10 chained aminoacids. Such peptides acquire physicochemical properties fitting the requirements of the downstream analytical methods, mainly the mass spectrometer (MS), which performs the data acquisition. The bottom-up paradigm is most often used because peptides are much more suitable to analysis by mass spectrometry, as illustrated in section \ref{subsec:the_detector}. This review will focus on the Data-Dependent Aquisition (\ac{DDA} approach within bottom-up proteomics, which currently is the most frequent workflow in proteomics \cite{Sinitcyn2018}. In this approach, the equipment is configured for the analysis of one peptide at a time, with the drawback of being biased for the most abundant peptides.

@@

An \ac{MS} experiment starts with the generation of a protein sample from the biological system of interest. Proteins are then denaturalized  so as to remove bias due to the divergent properties acquired by folded proteins. Then, proteins are subjected to digestion with specific proteases i.e protein-cutting molecules, which cut the aminoacidic chains following a predictable pattern. Trypsin is the most frequently used protease. It cuts peptidic bonds whenever a positively charged residue, either Lysine (K) or Arginine (R), lies on the carboxyl side of the peptidic bond. Since roughly 1/10 residues are either R or K, the average peptide length is 10 residues, as mentioned before. As demonstrated in section \ref{subsec:the_detector}, this length distribution is fitted to the resolution of the MS analyzer. Moreover, in as much R and K are positively charged aminoacids (see figure \ref{fig:aminoacids}), the resulting peptide is guaranteed in most cases to be able to capture at least one charge, which is key in the \ac{MS} workflow as described in section \ref{sec:the_mass_spectrometer}. All of these properties combined, together with its low prize, makes Trypsin the protease of choice in this step for most cases.

@@

Missed cleavages can happen due to steric impediments or the presence of specific aminoacids that can weaken the enzyme\textquotesingle s function \cite{Siepen2007}. This is the case of Trypsin whenever the residue on the other side of the peptidic bond is Proline. Altogether, a variability is created in the cleavage process that, though limited, needs to be taken care of in downstream analysis, as it could introduce biases in peptide observability.

@@

If presented with the problem of analyzing a mixture of peptides, the capacities of mass spectrometers are easily overwhelmed by a too complex mixture, resulting in the analysis of only a minor part of the total protein of the sample \cite{Barsnes2008}. This can be surmounted by analyzing one peptide in the sample at a time. The required sample separation is achieved by High-Performance Liquid Cromatography (\ac{HPLC}) methods, like reverse phase chromatography (separating on hydrophobicity) and strong cation exchange chromatography (separating on isoelectric point) \cite{Barsnes2008}.

%CITE 4.2 computational methods.
During \ac{HPLC}, the peptide mix is loaded into a column containing a stationary and a solid phase. These phases create an environment where peptides interact differently based on their physico-chemical properties, set by the nature of the phases. The output of the column, called elute, will consist of subsets or fractions of peptides leaving the column at different retention times (\ac{RT}) i.e the amount of time passed before the peptide is observed in the mass spectrometer. Therefore, the input to the spectrometer will consist of few peptides at any given time.

@@

The mass spectrometer is the ensemble of pieces of equipment analyzing a peptide like as those generated following the workflow enunciated in sections \ref{sec:sample_processing} and \ref{subsec:peptide_separation}. It consists of three main parts: an ion source, a mass analyzer, and a detector (see figure \ref{fig:mass_spectrometer}) \cite{Barsnes2008}.

@@

This step is performed in the ion source \cite{Barsnes2008}. The most frequent ionization methods in proteomics are Matrix-Assisted Laser Desorption-Ionization (\ac{MALDI}) and Electro Spray Ionization (\ac{ESI}) \cite{Mirzaei2016}. Most peptides ionized by MALDI will acquire a single charge, whereas ESI can provide multiple charges (+2, +3, etc) too \cite{Joshi2016}. Thus, the charge exhibited by an ion is not obvious when produced via ESI. Moreover, ESI can be run online with the right robotic equipment, while MALDI demands waiting time for vacuum generation. Finally, due to the chemical nature of the matrix components, MALDI ionizes more easily peptides containing aminoacids featuring aromatic rings (PYW) \cite{Hessling2013}, thus introducing a bias \cite{Hessling2013}. Bias in ESI is less predictable. All the sources of bias introduced during ionization cause the competitive ionization problem \cite{Zhang2009} \cite{Tang2004}.

The acquired charge yields a mass/charge (\ac{m/z}) ratio, a property that is applied in the downstream component separation and measurement steps. It is measured in $\frac{Da}{e}$, sometimes referred to as Th (Thomson).

@@

On the other hand, due to the natural occurrence of isotopes, particularly \ce{^{13}_{}C}, the same peptide will induce the measurement of several signals with very close \ac{m/z} values. They constitute the isotopic envelope of the ion (see figure \ref{fig:envelope}) \cite{Mirzaei2016}, and represent the signal created by peptides containing an increasing number of \ce{^{13}_{}C} atoms or other naturally occurring isotopes. Every time a \ce{^{12}_{}C} is replaced by \ce{^{13}_{}C}, the mass increases by 1 Da. Even though the natural abundance of \ce{^{13}_{}C} is 1.1 \%, the sheer number of carbon atoms in a peptide makes it likely that at least one or even more carbon atoms will be \ce{^{13}_{}C}, eventually driving the pure \ce{^{12}_{}C} signal to comparatively small intensity values, and down to intensities below the background noise. Such event can be problematic if it entails that the \ce{^{13}_{}C} peak is confused for the \ce{^{12}_{}C} peak.

@@

The resolution achieved by modern equipment allows for the distinction of each individual signal in most isotopic envelopes. Remarkably, the separation across peaks in the envelope can be used to infer the charge of the peptide with the following expression.

\begin{equation}
m/z = \frac{(m + z × H^+)}{z} \text{for \textit{z}} \in {1, 2,...}
\end{equation}

where $H^+$ is the mass of a single charge (1 Da) and $z$ is the number of charges acquired. A single charge will induce a separation of 1 \ac{m/z}, while at charge 2 it will be $1/2 = 0.5$ \ac{m/z}, at 3 $1/3 = 0.33$ \ac{m/z}, and so on (see figure \ref{fig:envelope}).

@@

Shotgun proteomics analyses make use of two or more mass spectrometers connected in series, giving rise to the so-called Tandem MS (\ac{MS/MS}) workflow. In this setting, each mass spectrometer collects a different type of spectra and thus different information (see figure \ref{fig:shotgun}). An extra spectrometer, usually a quadrupole, is introduced in between.

@@

\item The first spectrometer records the intensity versus \ac{m/z} ratio of the peptides eluting from the column at a given time and is used to filter ions exhibiting a selected \ac{m/z} ratio (in the \ac{DDA} protocol).

@@

\item The last spectrometer records the intensity versus \ac{m/z} ratio of the fragments produced in the previous step. The resulting spectrum can be use to read out the peptide sequence.

@@

Fragmentation occurs in the second spectrometer. But what is it, and why is it done? The information that can be extracted from \ac{MS1} scans consists of the \ac{m/z} ratio and retention time of the peptide ion, but not its sequence. Having the latter is paramount if the spectrum is to be matched to a theoretical peptide. Fortunately, the peptide sequence of the ion can be inferred if fragmentation is performed.

@@

During peptide fragmentation, bonds along the peptidic chain are broken, turning the peptide into smaller fragments. These fragments will consist of truncated versions of the original peptide at different positions, thus making it possible to read an \ac{m/z} ratio difference between any pair of fragment ions. The difference can be exploited to deduce which aminoacid makes up for that difference. If this process is repeated for enough pairs of contiguous fragments, with the right software, a sequence can be read from the \ac{MS/MS} spectrum, as  explained in section \ref{sec:search_engines}.

@@

Computational analysis of MS data starts with the matching of the spectra to a referene proteome (see figure \ref{fig:computational_analysis}). \ac{MS} search engines are capable of performing the crucial step of peptide to spectrum matching (PSM). During this step, search engines perform (I) \textit{in silico} prediction of the peptides produced in the protein digestion step and (II) their expected spectrum.

@@

Given the stochastic nature of the protein cleavage and spectra recording processes, the resulting data exhibit variability manifested in both missing and spurious peaks \cite{Stein1999}. Furthermore, random (wrong) matches can be returned by the PSM process when running against a sufficiently big database. This translates to the generation of multiple matches, of which one, if any, will be correct. Therefore, the lists of matches need to be somehow ranked by goodness-of-fit. The issue is addressed by search engines through the deployment of statistical models that provide scoring systems. Assuming the correct protein is present in the database, a good scoring system should give the best score to the right peptide. Under this circumstances, if repeated for several peptides, enough evidence for the presence of individual proteins can be collected.

Multiple search engines exist that implement different matching and scoring algorithms. The most modern ones include MS-GF+ \cite{Kim2014}, MS-Amanda \cite{Dorfer2014}, Comet \cite{Eng2013} and Andromeda \cite{Cox2011}. Notably, the results of each individual search engine can be combined to gather their strengths, at the expense of an increased computational cost and time \cite{Shteynberg2013}.

@@

A common filter is the false discovery rate (FDR), usually set to 1\%, indicating that after its application, only one out of a hundred filtered matches are expected to be false positives (wrong matches).

The most commonly used method to compute the \ac{FDR} of a list of matches is the target-decoy search. Using this method, the search engine replicates the matching process, using the same spectra, but instead against a decoy database i.e it simulates random matches. The decoy database is generated by reversing or more generally applying a randomization technique upon the sequences present in the original database (target) \cite{Elias2010}.

@@

The minimal FDR at which a given PSM is considered a valid match constitutes the PSM\textquotesingle s q-value, i.e it is the smallest FDR we can allow while still keeping the PSM \cite{Nesvizhskii2010}. Related to the q-value, the Posterior Error Probability (PEP) is an estimate of the probability of a given PSM of being an incorrect assignment (see figure \ref{fig:pep}). The PSM\textquotesingle s confidence is just defined as $1-PEP$ \cite{Nesvizhskii2010}. PEP can be computed from the decoy search results and provides another useful measurement of the uncertainty in the target results.

@@

The combination of all the aforementioned computational analyses yields a list of protein identities that reports the protein composition i.e qualitative information of the original sample. However, in most proteomics applications, quantitative data can be of great interest, as many biological phenomena are manifested mainly through changes in the protein abundances, cell compartments, rather than the simple dichotomy of the protein\textquotesingle s existence \cite{Barsnes2008}. For instance, cancer cells in response to a drug could modulate the abundance of several proteins without removing them from the cytosol or introducing new ones.

@@

Two paradigms exist in protein quantification: label-based and label-free. In label-based quantification, originally identical peptides from a number of different samples are made distinguishable by their masses via the incorporation of a label. All label-based methods simultaneously analyze several samples in each experiment, removing the difficulties associated with between-run variability \cite{Barsnes2008}. The finite number of "plexes" available for a given label sets the limit to how many samples can be differentially quantified \cite{Cox2014}. Different techniques, like Stable Isotope Labeling by Amino acids in Cell culture (SILAC) or Isotope-Coded Affinity Tags (ICAT), differ in the nature of the label and the way it is introduced. Most of them have potential limitations \cite{Patel2009}. This, together with the limited amount of samples that can be compared makes a case for label-free quantification.

@@

On the one hand, spectrum counting based quantification is the simplest method in proteomics. The number of peptides a peptide is detected in the spectrometer is detected as proxy for its abundance. It relies on the rationale that highly abundant peptides will have a higher intensity and are thus more likely to trigger the acquisition of MS/MS spectra. These methods have the advantage that they are very simple to implement and don't require any further data processing \cite{Barsnes2008}.

@@

For a more robust XIC based quantification, a feature extraction step is usually executed to extract consistent descriptors of identified peak clusters in the RT-m/z plance, as shown in figure \ref{figure:moff_apex}.

@@

It is for this reason that peptide-based models offer the statistical framework required to learn as much from the data as possible. This translates into improved results when compared to the other aforementioned methods \cite{Goeminne2015}. %This hypothesis is the motivation for the method explained in chapter \ref{chap:model}.

### 2018-07-25 pipeline.tex
Extend references

A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI \cite{Barsnes2018}, PeptideShaker \cite{Vaudel2015}, moFF \cite{Argentini2016} and MSqRob \cite{Goeminne2016} , was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software. The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) \ac{MBR} and feature extraction, (IV) protein inference and (V) relative quantification. Its output can be passed to follow-up analyses in R or Python to get a biological interpretation of the results.  A benchmark of its performance was accomplished using the proteome benchmark dataset published in \cite{Cox2014}. The results were comparable to those achieved by the MaxQuant \cite{Cox2014} software, excepting a bias produced by the sample fractionation of this dataset.

@@

Several proteomics pipelines are available on the internet under different licensing conditions. Many are released as closed-source commercial software, where information on how the program works is kept from the user. This is a serious drawback as it hinders the study of the implemented models and its customisation. Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) \cite{Deutsch2011} or openMS \cite{Sturm2008} are nevertheless available for the community, but they either lack good documentation, good broad base or format exchangeability. MaxQuant, a freeware but closed-source, monolithic proteomics analysis suite \cite{Cox2008}, is developed for Windows and has only very recently been ported to Linux \cite{Sinitcyn2018}. Nevertheless it has been extremely successfully adopted by the scientific community due to its ease of use and comprehensive pipeline.

@@

The preprocessing of the RAW files produced by the mass spectrometer into the MGF open format enabled searchGUI to dispose of the registered spectra. PeptideShaker quality control and filtering capabilities (see figure \ref{figure:qc_validation}) carried out the required search results validation. As expected, matches to the target and decoy exhibited similar score distributions at low score values, while a divergence is observed at higher score values. Likewise, the m/z error was found to be closer to 0 on validated PSMs than on those which did not pass the 1 \% FDR filter. The application of this filter implied that the FNR (false negative rate) was set to 5 \%, i.e 5 out of every 100 discarded matches were estimated to be true positives.

>>>>

A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI \cite{Barsnes2018}, PeptideShaker \cite{Vaudel2015}, moFF \cite{Argentini2016} and MSqRob \cite{Goeminne2016}, was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software. The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) protein inference, (IV) \ac{MBR} and feature extraction, and (V) relative quantification. Its output can be passed to follow-up analyses in online databases or Bioconductor to get a biological interpretation of the results.  A benchmark of its performance was accomplished using the proteome benchmark dataset published in \cite{Cox2014}. The results were comparable to those achieved by the MaxQuant \cite{Cox2014} software, excepting a bias produced by the sample fractionation of this dataset.

@@

Several proteomics pipelines are available on the internet under different licensing conditions. Many are released as closed-source commercial software, where information on how the program works is kept from the user. This is a serious drawback as it hinders the study of the implemented models and its customisation. Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) \cite{Deutsch2011} or openMS \cite{Sturm2008} are nevertheless available for the community, but they either lack good documentation, good broad base or format exchangeability. MaxQuant, a freeware but closed-source, monolithic proteomics analysis suite \cite{Cox2008}, is developed for Windows and has only very recently been ported to Linux \cite{Sinitcyn2018}. Nevertheless it has been extremely successfully adopted by the scientific community due to its ease of use and comprehensive pipeline. The organization of a pipeline with comprehensive documentation, open-source nature, and easy implemetation and customisation was asserted to be lacking in the literature.

@@

The preprocessing of the RAW files produced by the mass spectrometer into the MGF open format enabled searchGUI to dispose of the registered spectra. PeptideShaker quality control and filtering capabilities carried out the required search results validation. As expected, matches to the target and decoy exhibited similar score distributions at low score values, while a divergence is observed at higher score values (see figure \ref{figure:qc_validation}). Likewise, the m/z error was found to be closer to 0 on validated PSMs than on those which did not pass the 1 \% FDR filter. The application of this filter implied that the FNR (false negative rate) was set to 5 \%, i.e 5 out of every 100 discarded matches were estimated to be true positives.
